\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{listings}
\usepackage{syntax}
\usepackage{forest}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Design Specification: PARD Programming Language}
\author{Erdinç Daşkın}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

\section{Introduction \& Philosophy}
PARD is a programming language designed around game design and development.
It leverages FORTRAN's fast math and C interoperability.
The name PARD is an abbreviation of Punching Card, which is an homage to the early days of FORTRAN.
\section{Lexical Grammar}
\subsection{Variables}
\subsubsection{Variables and their types}
Variables, just like most other programming languages, are a significant part of PARD. \\
There's 4 basic variable types:
\begin{itemize}
    \item \textbf{int}: A standard integer type for counts, loop counters, and IDs.
    \item \textbf{float}: A floating-point number for positions, physics calculations, and deltatime.
    \item \textbf{bool}: A boolean value (\texttt{true} or \texttt{false}) for state flags.
    \item \textbf{string[N]}: A fixed-length, ultra-fast character array of size $N$.
\end{itemize}
As you can see, strings are not dynamically sized - this is not only for easy operation in fortran, but it also introduces a good performance boost.
\subsubsection{Variable definition}
You can easily define a variable in the following way: \\
\texttt{int myInt = 10} \\
\texttt{float myFloat = 10.5} \\
\texttt{bool myBool = true} \\ 
\texttt{string[6] myString = 'Hello!'}
\subsubsection{Simple usage of variables}
You can easily use a variable in multiple ways, here are some: \\
\texttt{int x = 2} \\
\texttt{int y = 3} \\
\texttt{int z = x + y} \\
\texttt{float f = 2.5} \\
\texttt{float g = 0.5} \\
\texttt{float h = f / g} \\
You cannot mix types: by this I mean you can't do the following: \\
\texttt{int x = 2} \\
\texttt{float y = 10.5} \\
\texttt{float z = x + y} \\
This will raise an error, as not all types used together are of the same type.
\subsubsection{Converting variables to other types}
In the previous example, we had this: \\
\texttt{int x = 2} \\
\texttt{float y = 10.5} \\
\texttt{float z = x + y} \\
But what if we want to combine different types? Well, you can use simple parsing functions:
\texttt{float z = x + y.tofloat} \\
There exists tofloat, toint and tostring.
\subsubsection{Array definition}
Okay, how do we define arrays? \\
Before the type, put (n) where n is the lenght of said array. \\
\texttt{(2)int myInt = [10, 5]} \\
\texttt{(3)float myFloat = [10.5, 3.14, 5.5]} \\
\texttt{(3)bool myBool = [true,false,false]} \\ 
\texttt{(2)string[6] myString = ['Hello ', 'World!']}
\subsubsection{Accessing values in an array}
To get a part of an array, use an index. PARD is 1-index based, e.g. the first item is index 1, the second is index 2 etc. \\
\texttt{(2)int myIntArray = [10, 5] \\ int myInt = myIntArray[2]} \\
In this exaple, myInt will be 5.
\subsubsection{Constants}
One can easily define constants using the const keyword. \\
\texttt{const (2)int myIntArray = [10, 5]} \\
\texttt{const float y = 10.5} \\
\subsection{Procedures}
One may ask, what is the difference between procedures and functions in PARD. Well, the difference is, that functions always return a value, while not all procedures have to. \\
All procedure parameters are passed by reference by default. \\
There's 3 types of procedures:
\subsubsection{Routines}
Routines are procedures that don't return any value. They are defined by the rout keyword and require to have atleast 1 parameter. \\
\texttt{rout myRoute(int x, int y)\{  \\ x = y \\ \}  } \\
Keep in mind, route parameters are by default passed by reference, and in this case, the original x passed will be modified.
\subsubsection{Subroutines}
Subroutines are routes that dont take any parameters. They are defined by the sub keyword. \\ 
\texttt{sub mySub \{\\  pass \\\}} \\
the pass keyword simply does nothing.
\subsubsection{Functions}
Functions are like routes that return a value. They are defined by the keyword fn. They may or may not have parameters. \\
\texttt{fn myFunc(int x, int y)\{  \\ return x\\ \}  } \\
\subsubsection{Passing parameters by value}
If you do not wish a routine/function to affect variables passed into it, \\you may use the  value indicator \&: \\
\texttt{rout myRoute(\&int x, int y)\{  \\ x = y \\ \}  } \\
In this case, if a variable is passed as the parameter x, it wont change to y globally.
\subsection{Engine procedures}
Now, to actually be able to function correctly and have a drawing loop, there are 3 predefined procedures by PARD which must be used in all programs. \\
\subsubsection{init}
The init sub is called when the program starts and its meant to initialize variables/state. \\
\texttt{sub init \{\\ pass \\\}}
The init sub should always be the first procedure in a file.
\subsubsection{draw}
The draw sub is called every game tick and is meant to draw stuff on the screen. \\
\texttt{sub draw \{\\ clear(WHITE) \\\}} \\
By the way, the clear routine clears the screen to the passed color.
\section{Compiler Architecture}
There are 2 big components in PARD in the compiling proccess: the actual PARD transpiler, and a fortran compiler.
The default compiler used by PARD is GFortran,
tho you can use your own compiler by using the -k (keep) flag during compiling, which makes it so only the PARD transpiler is used and the result is the .f90 file
\subsection{File extension}
PARD files don't require any specific extension (like most other languages), however, the .pard extension is highly recomended (to allow IDE functionality like intellisense etc.)
\subsection{The transpiling process}
PARD is not really a compiler, as it transpiles to FORTRAN.
It uses the raylib-c binding to have the game engine functionality (big thanks to Interkosmos btw)
Because FORTRAN is not case-sensitive, so is PARD, as all keywords are converted to lowercase during transpiling.
Because fortran doesnt allow for variable declaration mid-code, the transpiling proccess is 2pass:
\subsubsection{1st pass}
In the first pass, the transpiler scans the file for all variable declarations, and then puts them all at the start, and then continues to transpile.
\subsubsection{2nd pass}
In the second pass, the transpiler goes and transpiles the modified file into modern Fortran.
\end{document}